---
title: Natural Language Processing
author: Jeff Cavanagh
date: '2020-08-05'
slug: natural-language-processing
categories:
  - R
tags:
  - NLP
lastmod: '2020-08-05T15:49:17-04:00'
layout: post
type: post
highlight: yes
---

**Data is only numerical or categorial right?? **

Wrong. Not all data these days is numerical and easy to work with established mathetimatical techniques. Quite a bit of potential data is unstructured and textual in nature, which does not operate by the same rules as numerical data. However, even though it is formatted differently, there are still underlying patterns and ways to work to draw insight from this subset of data. In this article we will focus on some routine methods of Natural Language Processing (NLP) and how it can be utilized to process and draw value from text data. 

## Setup

### `stringr`

In R, the best way I have found to carry out any sort of NLP is using the `stringr` package. This package can be loaded on its own, or as part of `tidyverse`.

The functions contained with in all process character strings in various ways. Luckily for us, R has a [cheatsheet](/rcheatsheets/strings.pdf){target="blank"} for the `stringr` package for not only the functions, but also the formatting of search patterns.

It should be noted that there is also the `stringi` package which contains more robust functions for handling strings. For simplicity, we will only be concentrating on `stringr` so let's load that now.

```{r}
library(stringr)
```

### The data

To illustrate the use of NLP, we will be analyzing a pdf of Mary Shelley's, *Frankenstein* (downloaded from [planetbook.com](https://www.planetebook.com/free-ebooks/frankenstein.pdf){target="blank"}). We will use the `pdftools` package to extract the text from the pdf file.

```{r}
library(pdftools)
book <- pdf_text("frankenstein.pdf")
```

`pdf_tools`
