---
title: Hiking Trail Analysis
author: Jeff Cavanagh
date: '2019-12-12'
slug: hiking-trail-analysis
categories:
  - R
tags:
  - ML
lastmod: '2020-08-01T09:50:05-04:00'
layout: post
type: post
highlight: no
---

```{r install, eval = FALSE,echo=F}

## This code will not be evaluated automatically.
## (Notice the eval = FALSE declaration in the options section of the
## code chunk)

my_packages <- c("tidyverse", "broom", "coefplot", "cowplot",
                 "gapminder", "GGally", "ggrepel", "ggridges", "gridExtra",
                 "here", "interplot", "margins", "maps", "mapproj",
                 "mapdata", "MASS", "quantreg", "rlang", "scales",
                 "survey", "srvyr", "viridis", "viridisLite", "devtools")

install.packages(my_packages, repos = "http://cran.rstudio.com")

```
**The following was my final project for a Predictive Analytics course. It analyzed data on hiking trails throughout the US and created and compared models predicting whether or not a given trail will be rated above 4.5 stars out of 5.**

### Setup
```{r}
trails=read.csv("AllTrailsdata_nationalpark.csv",header=T)
attach(trails)
dim(trails)
```

This data set gives information on trails in the National Parks Service. The data is publicly available on Kaggle.com.

Each row is a different trail. The columns describing each trail are:

- trail_id
- name
- area_name
- city_name
- state_name
- country_name
- _geoloc (coordinates)
- popularity (scaled value)
- length (km)
- elevation_gain
- difficulty_rating (1:5)
- route_type (out and back, loop, point to point)
- visitor_usage (1:4)
- avg_rating (1:5)
- num_reviews
- features (dogs-no, forest, views, lake, wild-flowers, etc...)
- activities (birding, hiking, nature-trips, etc...)
- units (i,m)

Average rating was chosen as the dependent variable. The quantitative variables of popularity, length, elevation gain, difficulty rating, and number of reviews were then used as predictors in the classification problem of creating a model to accurately predict the average rating a trail will receive. 

```{r}
mean(avg_rating)
avg_rating_ideal=(avg_rating>4.5)*1
avg_rating_ideal=as.factor(avg_rating_ideal)

```

The mean value of average trail ratings was 4.17. To predict the highest tier of terms, the response variable average rating was turned into a classification variable with levels: avg_rating > 4.5, and avg_rating ≤ 4.5.

### Partitioning the Data
```{r}
set.seed(1)
train=sample(3313,0.7*3313)
trails_reg=cbind(popularity,length,elevation_gain,difficulty_rating,num_reviews,avg_rating_ideal)
trails_reg=data.frame(trails_reg)
training=trails_reg[train,]
testing=trails_reg[-train,]

```

### Interaction of Predictor Variables
```{r,echo=F,warning=F}
library(GGally)
ggpairs(trails_reg,upper=list(combo='box'), lower = list(continuous = wrap("smooth", method = "loess"), combo='facetdensity') )
```

### Model Building
#### Linear Regression
```{r}
lm.fit=lm(avg_rating_ideal~.,data=training)
lm.pred=predict(lm.fit,testing)
mean(abs(lm.pred-testing$avg_rating_ideal))
```

#### LDA
```{r}
library(mgcv)
library(MASS)
lda.fit=lda(avg_rating_ideal~.,data=training)
lda.pred=predict(lda.fit,testing)
table(testing$avg_rating_ideal,lda.pred$class)
mean(lda.pred$class!=testing$avg_rating_ideal)
```

#### QDA
```{r}
qda.fit=qda(avg_rating_ideal~.,data=training)
qda.pred=predict(qda.fit,testing)
table(testing$avg_rating_ideal,qda.pred$class)
mean(qda.pred$class!=testing$avg_rating_ideal)
```

#### Logistic Regression
```{r}
set.seed(1)
glm.fit=glm(as.factor(avg_rating_ideal)~.,data=training,family=binomial)
glm.probs=predict(glm.fit,testing,type='response')
glm.pred=rep("n",994)
glm.pred[glm.probs>0.5]="y"
set.seed(1)
table(testing$avg_rating_ideal,glm.pred)
1-(832+19)/994
```
#### KNN
k=3
```{r}
set.seed(1)
library(class)
train.X=cbind(popularity,length,elevation_gain,difficulty_rating,num_reviews)[train,]
test.X=cbind(popularity,length,elevation_gain,difficulty_rating,num_reviews)[-train,]
train.avg_rating_ideal=avg_rating_ideal[train]
knn3.pred=knn(train.X,test.X,train.avg_rating_ideal,k=3)
table(knn3.pred,testing$avg_rating_ideal)
1-(773+23)/994
```

k=4
```{r}
set.seed(1)
knn4.pred=knn(train.X,test.X,train.avg_rating_ideal,k=4)
table(knn4.pred,testing$avg_rating_ideal)
1-(770+27)/994
```

k=5
```{r}
set.seed(1)
knn5.pred=knn(train.X,test.X,train.avg_rating_ideal,k=5)
table(knn5.pred,testing$avg_rating_ideal)
1-(799+17)/994
```

k=6
```{r}
set.seed(1)
knn6.pred=knn(train.X,test.X,train.avg_rating_ideal,k=6)
table(knn6.pred,testing$avg_rating_ideal)
1-(799+21)/994
```

#### Cross Validation
```{r}
trails_reg=data.frame(trails_reg)
MAE=c()

# LM
mae=c()
for (i in 1:10){
     train_data=trails_reg[-((i*231-230):(i*231)), ]
     test_data=trails_reg[((i*231-230):(i*231)), ]
     lm.fit=lm(avg_rating_ideal~.,data=train_data)
     lm.pred=predict(lm.fit,test_data)
     
     mae=c(mae, mean(abs(lm.pred-test_data$avg_rating_ideal)))
}



MAE=c(MAE, mean(mae))

#LDA

mae=c()
for (i in 1:10){
     train_data=trails_reg[-((i*231-230):(i*231)), ]
     test_data=trails_reg[((i*231-230):(i*231)), ]
     lda.fit=lda(avg_rating_ideal~.,data=train_data)
     lda.pred=predict(lda.fit,test_data)
     
     mae=c(mae, mean(lda.pred$class!=test_data$avg_rating_ideal))
}



MAE=c(MAE, mean(mae))

#QDA
mae=c()  

for (i in 1:10){
    train_data=trails_reg[-((i*231-230):(i*231)), ]
     test_data=trails_reg[((i*231-230):(i*231)), ]
     qda.fit=qda(avg_rating_ideal~.,data=train_data)
     qda.pred=predict(qda.fit,test_data)
     mae=c(mae, mean(qda.pred$class!=test_data$avg_rating_ideal))
   
}


MAE=c(MAE, mean(mae))


#Logistic Regression

mae=c()  

for (i in 1:10){
     train_data=trails_reg[-((i*231-230):(i*231)), ]
     test_data=trails_reg[((i*231-230):(i*231)), ]
     glm.fit=glm(as.factor(avg_rating_ideal)~.,data=train_data,family=binomial)
     glm.probs=predict(glm.fit,test_data,type="response")
     glm.pred=rep("n",231)
     glm.pred[glm.probs>.5]="y"
     set.seed(1)
     table(test_data$avg_rating_ideal,glm.pred)
     mae=c(mae, 1-(203+2)/231)
}


MAE=c(MAE, mean(mae))


#knn 3


mae=c() 

for (i in 1:10){
    train_data=trails_reg[-((i*231-230):(i*231)), ]
     test_data=trails_reg[((i*231-230):(i*231)), ]
     
     train.X=cbind(train_data$popularity,train_data$length,train_data$elevation_gain,train_data$difficulty_rating,train_data$num_reviews)
     test.X=cbind(test_data$popularity,test_data$length,test_data$elevation_gain,test_data$difficulty_rating,test_data$num_review)
     train.avg_rating_ideal=train_data$avg_rating_ideal
     knn3.pred=knn(train.X,test.X,train.avg_rating_ideal,k=3)
     mae=c(mae,mean(knn3.pred!=test_data$avg_rating_ideal))
    
}


MAE=c(MAE, mean(mae))


#knn 4


mae=c() 

for (i in 1:10){
      train_data=trails_reg[-((i*231-230):(i*231)), ]
     test_data=trails_reg[((i*231-230):(i*231)), ]
    
      train.X=cbind(train_data$popularity,train_data$length,train_data$elevation_gain,train_data$difficulty_rating,train_data$num_reviews)
     test.X=cbind(test_data$popularity,test_data$length,test_data$elevation_gain,test_data$difficulty_rating,test_data$num_review)
     train.avg_rating_ideal=train_data$avg_rating_ideal
     knn4.pred=knn(train.X,test.X,train.avg_rating_ideal,k=4)
     mae=c(mae,mean(knn4.pred!=test_data$avg_rating_ideal))
}


MAE=c(MAE, mean(mae))

#knn 5


mae=c() 

for (i in 1:10){
      train_data=trails_reg[-((i*231-230):(i*231)), ]
     test_data=trails_reg[((i*231-230):(i*231)), ]
    
      train.X=cbind(train_data$popularity,train_data$length,train_data$elevation_gain,train_data$difficulty_rating,train_data$num_reviews)
     test.X=cbind(test_data$popularity,test_data$length,test_data$elevation_gain,test_data$difficulty_rating,test_data$num_review)
     train.avg_rating_ideal=train_data$avg_rating_ideal
     knn5.pred=knn(train.X,test.X,train.avg_rating_ideal,k=5)
     mae=c(mae,mean(knn5.pred!=test_data$avg_rating_ideal))
}

MAE=c(MAE, mean(mae))

#knn 6


mae=c() 

for (i in 1:10){
      train_data=trails_reg[-((i*231-230):(i*231)), ]
     test_data=trails_reg[((i*231-230):(i*231)), ]
    
      train.X=cbind(train_data$popularity,train_data$length,train_data$elevation_gain,train_data$difficulty_rating,train_data$num_reviews)
     test.X=cbind(test_data$popularity,test_data$length,test_data$elevation_gain,test_data$difficulty_rating,test_data$num_review)
     train.avg_rating_ideal=train_data$avg_rating_ideal
     knn6.pred=knn(train.X,test.X,train.avg_rating_ideal,k=6)
     mae=c(mae,mean(knn6.pred!=test_data$avg_rating_ideal))
}


MAE=c(MAE, mean(mae))


MAE
```

#### Graphical Comparison of Classification Error
```{r}
par(mfrow=c(1,2))

plot(MAE,xlim=c(0,9), col='blue', type='p', pch=19, main='Cross Validation for Trails Data', ylab='mean absolute error')

text(1:8, MAE, c('lm','lda', 'qda', 'logistic', 'knn3', 'knn4', 'knn5', 'knn6', ''), pos=2,  cex=0.5, col="red")


plot(MAE, xlim=c(0,9),ylim=c(0,1) , col='blue', type='p', pch=19, main='Cross Validation for Trails Data', ylab='mean absolute error')

text(1:8, MAE, c('lm','lda', 'qda', 'logistic', 'knn3', 'knn4', 'knn5', 'knn6'), pos=3,  cex=0.5, col="red")
```
### Bagging
```{r}
library(randomForest)

bagged_rating <- randomForest(as.factor(avg_rating_ideal)~.,data=training, mtry=5, importance =TRUE)

bagged_rating

```
```{r}
pre_bag = predict(bagged_rating, newdata = testing)
mean(pre_bag!=testing$avg_rating_ideal)

```

```{r}
bagged_rating <- randomForest(as.factor(avg_rating_ideal)~.,data=training, mtry=5, importance =TRUE, ntree=5000)
pre_bag = predict(bagged_rating, newdata = testing)
mean(pre_bag!=testing$avg_rating_ideal)
```

```{r}
library(caret)
importance(bagged_rating)
ratingimp=varImp(bagged_rating,scale=F)
plot(ratingimp,top=5)
varImpPlot(bagged_rating)

```

```{r}
library(caret)

ctrl <- trainControl(method = "cv",  number = 10) 

bagged_caret <- train(
  as.factor(avg_rating_ideal)~.,
  data = training,
  method = "treebag",
  trControl = ctrl,
  importance = TRUE
  )
bagged_caret
```

```{r}
pre_rating=predict(bagged_caret, newdata=testing)
mean(pre_rating!=testing$avg_rating_ideal)

```

Variable importance by caret. 
```{r}
ratingImp <- varImp(bagged_caret, scale = FALSE)
ratingImp
plot(ratingImp,top=5)
```
```{r}
bagged_rating <- randomForest(as.factor(avg_rating_ideal)~.,data=training, mtry=5, importance =TRUE)


ylm=max(bagged_rating$err.rate)

plot(bagged_rating$err.rate[,1], col='black', type = "l", ylim=c(0, ylm))
points(1:length(bagged_rating$err.rate[,2]), bagged_rating$err.rate[,2], type = "l", 
       col = "blue")
points(1:length(bagged_rating$err.rate[,3]), bagged_rating$err.rate[,3], type = "l", 
       col = "red")

plot(bagged_rating)

```

```{r}
pre_bag = predict(bagged_rating, newdata = testing)

err_bagg=mean(pre_bag!=testing$avg_rating_ideal)
err_bagg

MAE=c(MAE,err_bagg)
```

### GBM Classification

```{r}
library(caret)

fitControl <- trainControl(
                           method = "repeatedcv", 
                           number = 10,
                           repeats = 10 
                           )


```

```{r}
gbmFit1 <- train(as.factor(avg_rating_ideal) ~ ., data = training, 
                 method = "gbm", 
                 trControl = fitControl,
                 verbose = FALSE)
gbmFit1
(gbmFit1$finalModel)$tuneValue
```

```{r}


gbmGrid <-  expand.grid(interaction.depth = c(1, 5, 9), 
                        n.trees = (1:10)*50, 
                        shrinkage = c(0.001,0.01,0.1),
                        n.minobsinnode = c(5,10,20))
                        
nrow(gbmGrid)

gbmFit2 <- train(as.factor(avg_rating_ideal) ~ ., data = training, 
                 method = "gbm", 
                 trControl = fitControl, 
                 verbose = FALSE, 
                 tuneGrid = gbmGrid)
gbmFit2
(gbmFit2$finalModel)$tuneValue
```

```{r}
library(GGally)
plot(gbmFit1) 
plot(gbmFit2) 

```
#### Predict
```{r}
set.seed(1)
pred=predict(gbmFit1, testing)

table(pred, testing$avg_rating_ideal)

err_gbm1=1-(824+25)/994        #error rate
err_gbm1

MAE=c(MAE,err_gbm1)

pred=predict(gbmFit2, testing)

table(pred, testing$avg_rating_ideal)
err_gbm2=1-(830+26)/994        #error rate
err_gbm2

MAE=c(MAE,err_gbm1)
```

```{r}
par(mfrow=c(1,2))

plot(MAE,xlim=c(0,12), col='blue', type='p', pch=19, main='Error Comparison for Trails Data', ylab='mean absolute error')

text(1:11, MAE, c('lm','lda', 'qda', 'logistic', 'knn3', 'knn4', 'knn5', 'knn6', 'bagg','gbm1','gbm2'), pos=2,  cex=0.5, col="red")


plot(MAE, xlim=c(0,12),ylim=c(0,1) , col='blue', type='p', pch=19, main='Error Comparison for Trails Data', ylab='mean absolute error')

text(1:11, MAE, c('lm','lda', 'qda', 'logistic', 'knn3', 'knn4', 'knn5', 'knn6', 'bagg','gbm1','gbm2'), pos=3,  cex=0.5, col="red")
```